On-the-fly Algorithms for GKAT Equivalences \thanks{Identify applicable funding agency here. If none, delete this.}

### 2025-01-22 main.tex
small grammer fixes in first section

\author{\IEEEauthorblockN{Cheng Zhang\textsuperscript{\textsection}}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University College London}\\
London, United Kingdom \\
0000-0002-8197-6181}
\and
\IEEEauthorblockN{Qiancheng Fu}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
qcfu@bu.edu}
\and
\IEEEauthorblockN{Hang Ji\textsuperscript{\textsection}}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Cornell University}\\
Ithaca, USA \\
hj476@cornell.edu}
\and
\IEEEauthorblockN{Ines Santacruz}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
email address or ORCID}
\and
\IEEEauthorblockN{Marco Gaboardi}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
email address or ORCID}
}

@@

The lack of symbolic algorithms also makes GKAT a less desirable foundation for performance-critical applications.
Although the generated coalgebra from GKAT expression have a linear number of states, the number of transitions is exponential to \emph{all} the primitive tests appearing in the entire expression.
This inefficiency is noted in previous literature, and poses a real limitation on real-world systems~\cite{smolka_GuardedKleeneAlgebra_2020,zhang_CFGKATEfficientValidation_2025}.
On the other hand, symbolic algorithms for KAT have been proven to be effective for generic equality solving and domain-specific applications~\cite{pous_SymbolicAlgorithmsLanguage_2015,moeller_KATchFastSymbolic_2024}.

@@

One particularly important sub-coalgebra of a coalgebra \(S\) is the least sub-coalgebra that contains a state \(s\). 
We will denote this sub-coalgebra as \(⟨s⟩_{S}\), and call it \emph{principle sub-coalgebra} generated by \(s\). 
We sometimes omit the subscript \(S\) when it can be inferred from context or irrelevant.
Intuitively, we usually think of principle sub-coalgebra \(⟨s⟩_S\) as the sub-coalgebra that is formed by all the ``reachable state'' from \(s\).
This coalgebraic characterization of reachable state can allow us to avoid induction on the length of path from \(s\) to a state in \(⟨s⟩_S\).

@@

Guarded Kleene Algebra with Tests, or GKAT~\cite{smolka_GuardedKleeneAlgebra_2020}, is a deterministic fragment of Kleene Algebra with Tests. 
The syntax of GKAT over a set of primitive actions \(K\) and a set of primitive tests \(T\) can be defined in two sorts, boolean expressions \(\BExp\) and GKAT expressions \(\Exp\):
\begin{align*}
    a, b, c ∈ \BExp 
        & ≜ 1 ∣ 0 ∣ t ∈ T ∣ b ∧ c ∣ b ∨ c ∣ \overline{b} \\  
    e, f ∈ \Exp 
        & ≜ p ∈ K ∣ b ∈ \BExp ∣ e +_b f ∣ e ; f ∣ e^{(b)} 
\end{align*}
where \(e +_b f\) is the if statement with condition \(b\), \(e;f\) is the sequencing of expression \(e\) and \(f\), and \(e^{(b)}\) is the while loop with body \(e\) and condition \(b\).
We use notation like \(b ≤ c\), \(b ≡ c\), and \(b ≢ c\) for the usual order, equivalence, and inequivalence in Boolean Algebra.
A GKAT expression can be unfolded into a KAT expression in the usual manner~\cite{kozen_KleeneAlgebraTests_1997c}:
\begin{align*}
    e +_b f & ≜ b; e + \overline{b}; f &
    e^{(b)} & ≜ (b; e)^*; \overline{b}.
\end{align*}
Then the semantics of each expression \(⟦e⟧\) can be computed by the semantics of Kleene Algebra with Tests~\cite{kozen_KleeneAlgebraTests_1997c}.
An important construct in the semantics is \emph{atoms}, which are conjunctions of all the primitive tests either in its positive or negative form: for \(T ≜ \{t₁, t₂, …, tₙ\}\)
\[\At_T ≜ \{t₁' ∧ t₂' ∧ ⋯ ∧ tₙ' ∣ tᵢ' ∈ \{tᵢ, \overline{tᵢ}\}\}.\]
Follow the conventional notation, we denote an atom using \(α, β\); and we sometimes omit the subscript \(T\) when no confusing can arise.
Alternatively, atoms can also be thought of as truth assignments to each primitive tests, indicating which primitive tests is satisfied in the current program states; and \(α ≤ b\) if and only if the truth assignment represented by \(α\) satisfies \(b\).

>>>>

\author{\IEEEauthorblockN{Cheng Zhang\textsuperscript{\textsection}}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University College London}\\
London, United Kingdom \\
0000-0002-8197-6181}
\and
\IEEEauthorblockN{Qiancheng Fu}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
qcfu@bu.edu}
\and
\IEEEauthorblockN{Hang Ji\textsuperscript{\textsection}}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Cornell University}\\
Ithaca, USA \\
hj476@cornell.edu}
\and
\IEEEauthorblockN{Ines Santacruz}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
email address or ORCID}
\and
\IEEEauthorblockN{Marco Gaboardi}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
inessdv15@gmail.com}
}

@@

Secondly, the lack of symbolic algorithms makes GKAT a less desirable foundation for performance-critical applications.
Although the generated coalgebra from GKAT expression have a linear number of states, the number of transitions is exponential to \emph{all} the primitive tests appearing in the entire expression.
This inefficiency is noted in previous literature, and poses a real limitation on real-world systems~\cite{smolka_GuardedKleeneAlgebra_2020,zhang_CFGKATEfficientValidation_2025}.
On the other hand, symbolic algorithms for KAT have been proven to be effective for generic equality solving and domain-specific applications~\cite{pous_SymbolicAlgorithmsLanguage_2015,moeller_KATchFastSymbolic_2024}.

@@

One particularly important sub-coalgebra of a coalgebra \(S\) is the least sub-coalgebra that contains a state \(s\). 
We will denote this sub-coalgebra as \(⟨s⟩_{S}\), and call it \emph{principle sub-coalgebra} generated by \(s\). 
We sometimes omit the subscript \(S\) when it can be inferred from context or irrelevant.
Intuitively, we usually think of principle sub-coalgebra \(⟨s⟩_S\) as the sub-coalgebra that is formed by all the ``reachable state'' from \(s\).
This coalgebraic characterization of reachable state can allow us to avoid induction on the length of a path from \(s\) to a state in \(⟨s⟩_S\).

@@

Guarded Kleene Algebra with Tests, or GKAT~\cite{smolka_GuardedKleeneAlgebra_2020}, is a deterministic fragment of Kleene Algebra with Tests. 
The syntax of GKAT over a set of primitive actions \(K\) and a set of primitive tests \(T\) can be defined in two sorts, boolean expressions \(\BExp\) and GKAT expressions \(\Exp\):
\begin{align*}
    a, b, c ∈ \BExp 
        & ≜ 1 ∣ 0 ∣ t ∈ T ∣ b ∧ c ∣ b ∨ c ∣ \overline{b} \\  
    e, f ∈ \Exp 
        & ≜ p ∈ K ∣ b ∈ \BExp ∣ e +_b f ∣ e ; f ∣ e^{(b)} 
\end{align*}
where \(e +_b f\) is the if statement with condition \(b\), \(e;f\) is the sequencing of expression \(e\) and \(f\), and \(e^{(b)}\) is the while loop with body \(e\) and condition \(b\).
We use notation like \(b ≤ c\), \(b ≡ c\), and \(b ≢ c\) for the usual order, equivalence, and inequivalence in Boolean Algebra.
A GKAT expression can be unfolded into a KAT expression in the usual manner~\cite{kozen_KleeneAlgebraTests_1997c}:
\begin{align*}
    e +_b f & ≜ b; e + \overline{b}; f &
    e^{(b)} & ≜ (b; e)^*; \overline{b}.
\end{align*}
Then the semantics of each expression \(⟦e⟧\) can be computed by the semantics of Kleene Algebra with Tests~\cite{kozen_KleeneAlgebraTests_1997c}.
An important construct in the semantics is \emph{atoms}, which are conjunctions of all the primitive tests either in its positive or negative form: for \(T ≜ \{t₁, t₂, …, tₙ\}\)
\[\At_T ≜ \{t₁' ∧ t₂' ∧ ⋯ ∧ tₙ' ∣ tᵢ' ∈ \{tᵢ, \overline{tᵢ}\}\}.\]
Following the conventional notation, we denote an atom using \(α, β\); and we sometimes omit the subscript \(T\) when no confusion can arise.
Alternatively, atoms can also be thought of as truth assignments to each primitive tests, indicating which primitive tests is satisfied in the current program states; and \(α ≤ b\) if and only if the truth assignment represented by \(α\) satisfies \(b\).
